---
title: "Probability distributions"
output:
  html_document:
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
---

Probability Distributions
========================================================

Probability distributions are one of the building blocks of statistics and
come in handy for modeling undertain quantities when we build simulation models.
R has numerous functions for:

* generating random numbers from a wide range of probability distributions
* calculating probabilities from pdf's and CDF's
* calculating quantiles for a wide range of probability distributions

There's a good introduction in Ch 17 of **RforE**.

Generating random variates
---------------------------

Start with the familiar normal distribution
```{r gen_normals}
# Generate 10 draws for a N(0,1)
rnorm(n=10)

# Generate 25 draws for a N(10,3)
rnorm(n=25,10,3)
```

One reason for knowing about random number generation is that you'll see people
on StackOverflow use functions like `rnorm` and `runif` (which at first glance
suggests "run if" - nope) to generate simple random datasets to use in their
questions. For example, to make a quick vector of numbers:

```{r random_vector}
v <- rnorm(10)
v
```

Or, to create a dataframe:

```{r random_df}
d <- data.frame(a=rnorm(10),b=rnorm(10,5,1),c=runif(10,5,15))
str(d)
```

Probabilities and quantiles
----------------------------
You probably recall a few quantiles of the standard normal distribution.

Let $X \sim N(0,1)$. Find $x$ such that $P(X \leq x)=0.95$.

```{r qnorm}
qnorm(0.95,0,1)
```

What is $P(X \leq 3)$?

```{r pnorm}
pnorm(3,0,1)
```

How much area in the standard normal distribution is between +-2?

```{r norm_area_between}
pnorm(2,0,1) - pnorm(-2,0,1)
```

What about +-1, +-2, +-3?

```{r empirical_rule}
pnorm(1,0,1) - pnorm(-1,0,1)
pnorm(2,0,1) - pnorm(-2,0,1)
pnorm(3,0,1) - pnorm(-3,0,1)
```

We can plot the normal distribution density function.

```{r lib_ggplot2}
library(ggplot2)
```

```{r plot_norm_density}
seqNorm <- seq(-5,5,.001)
randDensity <- dnorm(seqNorm)
ggplot(data.frame(x=seqNorm, y=randDensity)) + 
  aes(x=x, y=y) + geom_line() + labs(x="Z values", y="Density")
```

This is just a quick first look at probability distributions in R. It's got a slew of functions. 